---
title: "Tutorial_Orca_matrix_plot_corr"
author: "Nicolas MARY"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
      number_sections: yes
      toc: yes
      toc_depth: 3
      code_folding: show
---

# Package install

This tutorial requires several packages. All packages are either available on Cran or Bioconductor. Only 1 package created specifically for this tutorial can be installed from GitHub.

First of all, to install packages from GitHub we need to install some packages:
```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biocViews")
install.packages("devtools")
```

Now you should be able to install TADkit_dev2 from GitHub with:
```{r, eval=FALSE}
devtools::install_github("Nico-FR/TADkit_dev2")
```

To learn what you can do with this package and how to do it, you should follow this tutorial:
https://github.com/Nico-FR/TADkit

# Load packages

Here are the packages we need. If they are not installed on your computer, you must install them first.
```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(TADkitdev2)
library(BiocGenerics)
```


# Control matrices

## Load 2 matrices

```{r}
#load control matrix of bovin "3654" at 64kb resolution for chr 1 as sparse matrix
mat3654_norm = read.table("./Control_matrices/mat3654_ice_chr1_64kb.txt", sep = " ", h = F) %>% as.matrix %>% as("CsparseMatrix")

#get Observed / Expected matrix
mat3654_norm_OE = matObsExp(mat3654_norm)

#idem for bovin "0197"
mat0197_norm = read.table("./Control_matrices/mat0197_ice_chr1_64kb.txt", sep = " ", h = F) %>% as.matrix %>% as("CsparseMatrix")
mat0197_norm_OE = matObsExp(mat0197_norm)
```

## Plot matrices

```{r}
#plot 1 matrix between 1 to 11Mb
MATplot(matrix = mat3654_norm,
        start = 8e6, stop = 24e6,
        bin.width = 64e3,
        log2 = TRUE,
        scale.color = "H",
        tad.upper.tri = "./Control_matrices/tad3654_10kb.bed")+
  ggtitle("matrix 3654 normalized")

MATplot(matrix = mat3654_norm_OE,
        start = 8e6, stop = 24e6,
        bin.width = 64e3,
        log2 = TRUE,
        scale.color = "OE",
        tad.upper.tri = "./Control_matrices/tad3654_10kb.bed")+
  ggtitle("Obs/Exp matrix 3654 normalized")

#plot 2 matrices: 3654 & 0197 (upper and lower part of the matrix respectively)
mMATplot(matrix.upper = mat3654_norm_OE,
         matrix.lower = mat0197_norm_OE,
         start = 8e6, stop = 24e6, bin.width = 64e3,
         log2 = TRUE, scale.colors = "OE",
         matrix.upper.txt = "3654",
         matrix.lower.txt = "0197",
         tad.upper.tri = "./Control_matrices/tad3654_10kb.bed",
         tad.lower.tri = "./Control_matrices/tad0197_10kb.bed")+
  ggtitle("Obs/Exp matrices normalized")
```

## Correlation analysis

### Observed

Create a list with the matrices
```{r}
#create a list with the 2 matrices:
mat_norm.lst = list(mat0197 = mat0197_norm,
                    mat3654 = mat3654_norm)
```

Compute Pearson correlation coefficient between observed counts of all counts and plot the graph:
```{r}
matCorr(
  matrice.lst = mat_norm.lst,
  log2 = FALSE,
  output = "plot",
  self_interaction = TRUE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")+
  ggtitle("observed counts")
```

This graph shows that the correlation is significantly impacted by bins with many interactions (top right of the graph). Most of these values come from the number of interactions within a bin (i.e. the diagonal of the matrix). So we can remove then with the parameter: self_interaction = FALSE.
```{r}
matCorr(
  matrice.lst = mat_norm.lst,
  log2 = FALSE,
  output = "plot",
  self_interaction = FALSE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")+
  ggtitle("observed counts (without diagonal)")
```

Here too, the correlation coefficient is mainly influenced by the few values with a large number of interactions. Knowing that the number of interactions decreases exponentially with distance you can use the log(observed count):
```{r}
matCorr(
  matrice.lst = mat_norm.lst,
  log2 = TRUE,
  output = "plot",
  self_interaction = FALSE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")+
  ggtitle("log(observed counts)")
```

### Observed /Expected

To avoid the effect of distance between interactions, we can use Observed / Expected matrices:
```{r}
#create a list with the 2 matrices:
mat_norm_OE.lst = list(mat0197 = mat0197_norm_OE,
                    mat3654 = mat3654_norm_OE)


matCorr(
  matrice.lst = mat_norm_OE.lst,
  log2 = TRUE,
  output = "plot",
  self_interaction = TRUE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")+
  ggtitle("log2(observed / Expected counts)")

```

Finally, this correlation can only be made for interactions for which distances (between bins) are less than 3.2Mb.
```{r}
matCorr(
  matrice.lst = mat_norm_OE.lst,
  log2 = TRUE,
  output = "plot",
  self_interaction = TRUE,
  max.distance = 3.2e6,
  bin.width = 64e3,
  method = "pearson")+
  ggtitle("log2(observed / Expected counts) ; interact. dist. <= 3.2Mb")
```


# Orca matrices

Orca output 250x250 log(observed/expected) and log(expected) matrices at different resolutions. Here's the table of possible Orca predictions :
```{r echo=FALSE}
data.frame(
  nb_bins = 250,
  bin_width = c(4e3, 8e3, 16e3, 32e3, 64e3, 128e3),
  predicton_size = c(4e3, 8e3, 16e3, 32e3, 64e3, 128e3) * 250
)
```

We'll therefore use the same resolution of the control matrices (i.e. 64kb) which give us the predicted matrices for a 16Mb sequence. 
```{r}
matOrca = read.table("./Orca/bos_taurus/wildtype/orca_predictions_16Mb.txt", header = FALSE) %>% as.matrix()

#plot the 250 bins
MATplot(matrix = matOrca,
        start = 1, stop = 16e6, bin.width = 64e3,
        scale.colors = "OE",
        log2 = FALSE)
```

## Load Orca matrices

As mentioned above, Orca output 250x250 log(observed/expected) and log(expected) matrices at different resolutions. On the other hand, our control matrices have a bin number that depends on the resolution. For a resolution of 64kb, the bin number of chromosome 1, which has a size of 158.53411 Mb, is 158.53411e6 / 64e3 = 2478.

To compare matrices, the simplest option is to create an empty matrix (of 2478 bins) into which we'll add Orca matrix at the right position (i.e. between 8 and 24Mb). To do this, we've created a function that returns the observed or observed/expected matrix:

### Observed / expected

```{r}
#load observed/expected matrix of size = 16Mb and centered at 16Mb
matOrca_OE = orca2matrix(
  df_prediction.path = "./Orca/bos_taurus/wildtype/orca_predictions_16Mb.txt", 
  sep = "\t",
  mpos = 16e6, 
  chromsize = 158.53411e6, 
  output = "OE", scale = 16e6)

TADkit::MATplot(matOrca_OE, bin.width = 64e3, start = 8e6, stop = 24e6, scale.colors = "OE", log2 = T)
```

### Observed

```{r}
#load observed matrix of size = 16Mb and centered at 2Mb
matOrca = orca2matrix(
  df_prediction.path = "./Orca/bos_taurus/wildtype/orca_predictions_16Mb.txt", 
  sep = "\t",
  mpos = 16e6, 
  chromsize = 158.534110e6, 
  output = "Obs", scale = 16e6,
  df_normmats.path = "./Orca/bos_taurus/wildtype/orca_normmats_16Mb.txt")

TADkit::MATplot(matOrca, bin.width = 64e3, start = 8e6, stop = 24e6, scale.colors = "H", log2 = T)
```

# Orca VS control matrices

```{r}
mMATplot(matrix.upper = matOrca_OE, matrix.lower = mat0197_norm_OE, bin.width = 64e3, start = 8e6, stop = 24e6, scale.colors = "OE", log2 = T, matrix.upper.txt = "orca", matrix.lower.txt = "0197")+ggtitle("log2(observed / Expected)")
```

## Fold changes between 2 matrices
```{r}
#fold changes between 2 matrices
MATplot(matrix = mat0197_norm_OE / matOrca_OE, bin.width = 64e3, start = 8e6, stop = 24e6, scale.colors = "OE2", log2 = T)+ggtitle("log2(mat0197_norm_OE / matOrca_OE)")
```

## Correlations analysis

### Observed

Create a list with all the matrices:
```{r}
mat_norm.lst = list(orca = matOrca,
                    mat0197 = mat0197_norm,
                    mat3654 = mat3654_norm)
```

Compute correlations between matrices:
```{r}
matCorr(
  matrice.lst = mat_norm.lst,
  log2 = TRUE,
  output = "corr",
  self_interaction = FALSE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")
```

Plot correlations between the first two matrices:
```{r}
matCorr(
  matrice.lst = mat_norm.lst,
  log2 = TRUE,
  output = "plot",
  self_interaction = FALSE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")+
  ggtitle("log2(observed counts)")
```

### Observed / Expected

#### All bins

```{r}
mat_norm_OE.lst = list(orca = matOrca_OE,
                    mat0197 = mat0197_norm_OE,
                    mat3654 = mat3654_norm_OE)

matCorr(
  matrice.lst = mat_norm_OE.lst,
  log2 = TRUE,
  output = "corr",
  self_interaction = TRUE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")

matCorr(
  matrice.lst = mat_norm_OE.lst,
  log2 = TRUE,
  output = "plot",
  self_interaction = TRUE,
  max.distance = NULL,
  bin.width = NULL,
  method = "pearson")+
  ggtitle("log2(observed / Expected)")
```


#### Bin distances <= 3.2Mb

```{r}
matCorr(
  matrice.lst = mat_norm_OE.lst,
  log2 = TRUE,
  output = "corr",
  self_interaction = TRUE,
  max.distance = 3.2e6,
  bin.width = 64e3,
  method = "pearson")

matCorr(
  matrice.lst = mat_norm_OE.lst,
  log2 = TRUE,
  output = "plot",
  self_interaction = TRUE,
  max.distance = 3.2e6,
  bin.width = 64e3,
  method = "pearson")+
  ggtitle("log2(observed / Expected)")
```



## Pileup of TAD boundaries

### Observed

```{r}
#load TAD domains as GRanges (.bed to GRanges)
tad0197.gr = dataframes2grange(
  annotation.table = read.table("./Control_matrices/tad3654_10kb.bed", h = F, sep = "\t"),
  chromsize = data.frame(chr = 1, size = 158.53411e6))

#filter TAD within Orca matrix
window = 1.6e6 # matrices (that we'll stack with MATfeatures function) = TAD starts +/-1.6Mb 
tad0197_8_24Mb.gr = tad0197.gr[start(tad0197.gr) >= 8e6 + window &
                                 start(tad0197.gr) <= 24e6 - window]

#pileup
MATfeatures(
  matrix = mat0197_norm,
  bin.width = 64e3,
  annot.gr = tad0197_8_24Mb.gr,
  chr = 1,
  window.size = window,
  output = "plot")+ggtitle("Observed matrix pileup of 20 TAD +/- 1.6Mb: 0197")

MATfeatures(
  matrix = matOrca,
  bin.width = 64e3,
  annot.gr = tad0197_8_24Mb.gr,
  chr = 1,
  window.size = window,
  output = "plot")+ggtitle("Observed matrix pileup of 20 TAD +/- 1.6Mb: Orca")
```

### Observed / expected

To compare orca matrices with matrices obtained differently (e.g. control matrices), it's best to use matrices that are normalized by distance (i.e. O/E). In this case, the MATfeature function cannot be used directly to produce the graph. Instead, we first need to use the MATfeature function to obtain the sum of the O/E matrices and then make the graph.

```{r}
pileup0197 = MATfeatures(
  matrix = mat0197_norm_OE,
  bin.width = 64e3,
  annot.gr = tad0197_8_24Mb.gr,
  chr = 1,
  window.size = window,
  output = "matrix")


pileupOrca = MATfeatures(
  matrix = matOrca_OE,
  bin.width = 64e3,
  annot.gr = tad0197_8_24Mb.gr,
  chr = 1,
  window.size = window,
  output = "matrix")


mMATplot(matrix.upper = pileup0197 / 20, # /20 to get mean(ObsExp) instead of sum(20 x ObsExp)
         matrix.lower = pileupOrca / 20,
         start = 1, stop = (window + 64e3) * 2, bin.width = 64e3,
         log2 = TRUE,
         scale.colors = "OE")
```

## view point

We'd now like to look at how one region (e.g. a TAD) interacts with other bins.
Let check the interactions of this TAD:

```{r}
#TAD for analysis:
tad = read.table("./Control_matrices/tad0197_10kb.bed", h=F, sep = "\t")[28,]

#plot 
mMATplot(matrix.upper = mat3654_norm_OE,
         matrix.lower = matOrca_OE,
        start = 8e6, stop = 20e6,
        bin.width = 64e3,
        log2 = TRUE,
        scale.color = "OE",
        tad.upper.tri = tad,
        tad.lower.tri = tad,
        matrix.upper.txt = "3654",
        matrix.lower.txt = "Orca")
```

We can visualize the interaction of this TAD with the rest of the genome with viewPointInteract funtion:

```{r}
viewPointInteract(matrix.lst = mat_norm.lst,
                  bin.width = 64e3,
                  vp.start = tad$V2 + 64e3, vp.stop = tad$V3 + 64e3,
                  start = 8e6, stop = 20e6,
                  log2 = T, norm = T)+ggtitle("VP interactions: Observed")

viewPointInteract(matrix.lst = mat_norm_OE.lst,
                  bin.width = 64e3,
                  vp.start = tad$V2 + 64e3, vp.stop = tad$V3 + 64e3,
                  start = 8e6, stop = 20e6,
                  log2 = T, norm = F)+ggtitle("VP interaction: Obs /Exp")

```






























